Exploring the Vectorization of Python Constructs Using Pythran and Boost SIMD

### 2013-11-29 doc/papers/wpmvp14/paper.tex
++Proof reading

    The Python language is highly dynamic, most notably due to late binding. As
    a consequence, program run using Python typically run an order of magnitude
    slower than their C counterpart. It is also a high level languages whose
    semantic can be made more static without much change from a user point of
    view in the case of mathematical applications. In that case, the language
    provides several vectorization opportunities that are studied in this
    paper, and evaluated in the context of Pythran, an ahead-of-time compiler
    that turns Python module into C++ meta-programs.

@@

For instance, the natural solution to Project Euler sixth problem ``Find the
difference between the sum of the squares of the first one hundred natural
numbers and the square of the sum`` can be coded as in
Listing~\ref{lst:euler06-py} in Python and as in Listing~\ref{lst:euler06-c} in
C. A comparison of the performance of the Python function and the C function
called through the \texttt{ctypes} module shows that the C version runs more
than $\times250$ faster than the Python version. Enabling compiler
auto-vectorization makes the C version $\times400$ faster than the Python
version. To perform more reliable time measurements, \texttt{n} has been
increased to 100001 instead of 101.

There are two main reason for the poor performance of Python:

@@

    \item[dynamic binding] each variable access is made through a dictionary
        look-up to bind variable names to variable instances~;

@@

In that context, it does not make sense to speak about vectorization: the
nature of an operation is unknown to the interpreter until its execution, and
data are not contiguous in memory.

@@

Several approaches have been proposed to solve the performance issues of
Python: using native modules, relying on Ahead-Of-Time (AOT) or Just-In-Time
(JIT) compilation.

A native module is a shared library whose interface make it callable from the
Python interpreter using the regular \texttt{import} mechanism. It makes it
possible to call native code from Python in a transparent way. Typically Python
object are unboxed when transferred from Python to C, where efficient
computations can happen. Then the result is boxed when getting back to the
Python world. In between, vectorization can happen. This is the approach taken
by the \texttt{numpy} module: a native type, the \texttt{ndarray} contains
contiguous unboxed elements, and provides a great deal of high-level functions
to manipulate them. All the computation-intensive part is done in pure C.
Vectorization can eventually happen at this level, for instance when performing
a point-to-point operation like the sum of two arrays. However the granularity
of the operations limits the ratio of \texttt{load}/\texttt{store} versus
operation. For instance, if \texttt{a}, \texttt{b} and \texttt{c} are 1-D
arrays, the sum \texttt{a + b + c} is computed using two loops and an
intermediate array.  It means there are 4 load / stores and only two additions
per loop iteration.

AOT compilation uses type inference or type annotation to statically compute
the type of all (like in Shedskin~\cite{shedskin2006},
Parakeet~\cite{parakeet2012}) or part of (like in Cython~\cite{cython2010} or
Numba~\cite{numba}) variables, effectively unboxing them to speed-up
computations. Using JIT compilation, type information is available at runtime
and the unboxing decision is based on profitability, e.g.\ according to a trace
analysis. This is the approach of the PyPy project~\cite{pypy2009}. In both
cases, generation of vector instruction is possible, but none of the cited
project use it. Indeed automatic vectorization is a complex topic, and as
stated in~\cite{maleki2011}, many cases are not yet correctly supported even
for mainstream compilers.

@@

Pythran~\cite{pythran2013, pyhpc2013} is also an AOT compiler, bit it holds the
unique property of turning Python modules into C++ \emph{meta-programs}. It
does not need type annotations and keeps the polymorphism of original
functions, respecting Python's duck typing. Its compilation workflow is
described in Figure~\ref{fig:pythran-compiler}.  Like others AOT compilers, it
unboxes all variables for better performance, and performs a wide variety of
optimizations such as lazy evaluation, constant folding, expression fusion etc.
It supports core \texttt{numpy} constructs and most Python constructs to the
notable exception of user classes. Constructs such as \texttt{eval}, are not
supported either.

@@

\caption{Pythran compiler workflow.}
\label{fig:pythran-compiler}
\end{figure}

This paper explores the vectorization of Python programs based on the idea that
many Python constructs already exhibit good vectorization opportunities. The
goal of the paper switches from automatic vectorization to vectorization of
high-level Python intrinsics, as described in Section~\ref{sec:python-semantic}.
One of the difficulties of vectorizing Python functions is linked to function
polymorphism. Vectorization in the context of C++ meta-programs is examined in
Section~\ref{sec:meta-vectorization}. The approach is validated using the
Pythran compiler on several kernels presented in Section~\ref{sec:benchs}.

@@

\boostsimd{} is a high-level C++ library to program SIMD architectures. 
Designed as an \textit{Embedded Domain Specific Language}, \boostsimd{} 
provides both expressiveness and performance by using generic programming 
to handle vectorization. \boostsimd{} does not only provide a portable 
way but also enable the use of common programming idioms when designing 
SIMD-aware algorithms. Two components are provided:

@@

    \item[portability] Only SSE was supported, moving to AVX or another vector
        instruction set would require a great deal of engineering;

    \item[completness] Many basic mathematic functions such as cos or sin
        are not directly available in the target instruction sets;

    \item[typing] Pytrhan manipulates meta-functions with parametric types,
        while a given instruction set only provide concretes type.

@@

Pythran uses function objects to represent any Python functions, either
user-functions or functions from the standard library. The former are
automatically generated from the user code analysis, and the latter are
hand-coded in a library called \texttt{pythonic++} provided with the compiler.
For instance the Python function \texttt{def foo(x): return 3 * x} is compiled
into a code similar to the one in Listing~\ref{lst:simple-pythran-output}. The
actual would be more complicated to handle type inference across several
statements, but the structure remains valid.

@@

This representation perfectly matches the requirements of \boostsimd, as
presented in Listing~\ref{lst:square-cxx}. The \texttt{foo} function can be
used indifferently on scalar types and vector types.

@@

A high-level language such as Python exhibits another level of vectorization
opportunity in implicit loop. Implicit loops can be found in many places, but
this paper focuses on \emph{list comprehension} and \emph{generator
expression}, \emph{map} or \emph{sum} intrinsics and \emph{array operations}
through the \texttt{numpy} module.

@@

The \texttt{sum} intrinsic is a typical case of potential vectorization of an
implicit loop. Its prototype is: \texttt{sum(sequence[, start]) $\rightarrow$
value} and its behavior is equivalent to the Python code in
Listing~\ref{lst:sum-py}. Depending on the type of \texttt{sequence}, and also
whether one needs to strictly conforms to the IEEE Standard for Floating-Point
Arithmetic (IEEE 754), vectorization of this reduction is a comm case of \texttt{fold}.

@@

Condition~\ref{enu:same-types} is a prerequiste for static typing and efficient
unboxing. For instance, any code that does not match it will fail to compile
using Pythran or Shed-Skin. It is also satisfied for the \texttt{ndarray}
container as they wraps unboxed arrays. Condition~\ref{enu:scalar-types} can be
checked at compile time. Python supports two native scalar types candidates for
vectorization: \texttt{int} (64 bits integer) and \texttt{float}
(double-precision floating point numbers). The \texttt{numpy} model add supports for a
wide range of signed and unsigned integers as well as single, double or
quadruple precision float. Depending on the target architecture, only some of
them can fit into vector registers. Finally condition~\ref{enu:contiguous} also
is a static property of a container. It is verified for C++ alternative to
Python's \texttt{list}, \texttt{str} and \texttt{ndarrays}, but generally not
for \texttt{set} or \texttt{dict}. Condition \ref{enu:random-access} is not
matched by \emph{generator}s, the Python incarnation of co-routines.

@@

As function have to be generic but some containers do not match required
properties, it is possible to select a vectorized
implementation of fall-back to a generic one using Substitution Failure
Is Not An Error (SFINAE)~\cite{metaprogramming2002}.

@@

\begin{lstlisting}[language=c++, label={lst:sum-cxx}, caption={Meta-implementation of the \texttt{sum} intrinsic in C++11.}]
template<class T, class V=long>
typename std::enable_if<is_vectorizable<T>::value,
                        typename T::value_type
                       >::type
sum(T const& sequence, V value=0L)
{
  // E is the type of an element of T
  typedef typename T::value_type E;
  // vE is a vector type of vsize elements of type E
  typedef typename simd::native<E,
    BOOST_SIMD_DEFAULT_EXTENSION> vE;
  static const size_t vsize =
    simd::meta::cardinal_of<vE>::value;

  // vector_value is for the parallel reduction
  vE vector_value = simd::splat<vE>(E());
  const size_t n = sequence.size();
  const size_t bound = n / vsize * vsize;
  for(size_t i = 0; i < bound; i+= vsize)
    vp += sequence.load(i);
  // inner sum of the vector
  value += boost::simd::sum(vp);

@@

Deciding if a given function is vectorizable can be approximated through a
composition rules. First, some functions are known to be vectorizable. This is
generally the case for the addition, subtraction etc. Depending on the target
vector instruction set, some operations may or may not be available as vector
instructions, e.g.\ trigonometric functions, but they can be provided as
composite instructions. Some are just not available, as quadruple precision
float addition. Starting from this set of vectorizable function, one can use
the \texttt{;} as a closure binary relation to compute a subset of all
vectorizable functions. Said otherwise, any sequence of operation that only
involve vector operations is a vector operation and a function whose body is a
sequence of vector operation or vectorizable functions is a vectorizable
function itself.

The preceding statement is only valid if the function's body only references
vector functions and formal parameters, as the function must be valid for both
scalar types and vector types. Note that this perfectly matches the translation
of polymorphic Python functions into C++ meta-functions, as detailed in
Listings~\ref{lst:square-python} and~\ref{lst:square-cxx}. The core idea is that
the \texttt{square} function can be called indifferently on floating point
values or on vector of floating point values. Through (automatic) template
instantiation, the C++ compiler generates the actual scalar and vector version
from the same meta-function. As Pythran always translates Python's polymorphic
function in that form, writing a vector version of the \texttt{map} intrinsic is
slightly more difficult than for the \texttt{sum} intrinsic due to its variadic
number of arguments, but it also relies on the \texttt{load}/\texttt{store}
protocol and SFINAE to select it.

@@

Generator expression is somehow similar to list comprehension, to the notable
exception that the value returned is an \emph{iterator adaptator} on the
container argument. In the context of vectorization, a generator expression
supports the vector protocol if the associated container also supports it. In
that case the  \emph{iterator adaptator} implements the vector protocol by
applying the comprehension rule to each vector element from this container, in a
streaming fashion.

Python also provide same features for set and dictionary.  Set comprehension
can be rewritten in the form of generator expression using the eponym function:
\texttt{\{ x + 3 for x in l\}} is equivalent to \texttt{set(x + 3 for x in l)}.
Dictionary comprehension and multi-level comprehension (i.e.\ containing
multiple \texttt{for} clauses) are not considered in this preliminary work.

@@

The \texttt{numpy} module played a major role in the adoption of Python by the
scientific community. In a nutshell, this is a thin Python wrapper above a raw
array that provides an array language embedded into Python. An example of
\texttt{numpy} code is show on Listing~\ref{lst:numpy-code}.

@@

Common features of array language include indexing, slicing, point-to-point
operation etc. The latter is traditionally a good source of vectorization.
However, because of the module layer, \texttt{numpy} always create a new array
to hold intermediate result. For instance in the expression \texttt{(1 -
x[:-1]) ** 2)} two array are created ---slicing does not create an array but a
view. The implied extra cost would hide the benefits of vectorization.

The typical way to solve this issue in C++ is to use \emph{expression
templates}, a technique described in~\cite{} that uses the C++ type system to
perform lazy evaluation of an expression. However, this approach is limited to
a single expression. Listing~\ref{lst:numpy-code-fs} illustrates the situation:
expression templates would not prevent the creation of temporary arrays
\texttt{t0} and \texttt{t1}. Fortunately, it is possible to apply \emph{forward
substitution} to fall-back to the expression from Listing~\ref{lst:numpy-code}
and avoid the creation of any temporary.

@@

It is possible to hook in the expression template to add vectorization support.
Indeed the expression from Listing~\ref{lst:numpy-code} hides a single vector
loop, that can be explicitly vectorized using \boostsimd in a similar manner as
the \texttt{map} and \texttt{sum} intrinsics.

@@

>>>>

    The Python language is highly dynamic, most notably due to late binding. As
    a consequence, programs run using Python typically run an order of magnitude
    slower than their C counterpart. It is also a high level language whose
    semantic can be made more static without much change from a user point of
    view in the case of mathematical applications. In that case, the language
    provides several vectorization opportunities that are studied in this
    paper, and evaluated in the context of Pythran, an ahead-of-time compiler
    that turns Python module into C++ meta-programs.

@@

To illustrate the relatively bad performance of Python for numerical
computations, let us consider the sixth problem from the Project Euler: ``Find
the difference between the sum of the squares of the first one hundred natural
numbers and the square of the sum``. A straight-forward solution can be coded
as in Listing~\ref{lst:euler06-py} in Python or as in
Listing~\ref{lst:euler06-c} in C. A comparison of the performance of the Python
function and the C function called through the \texttt{ctypes} module shows
that the C version runs more than $\times250$ faster than the Python version.
Enabling compiler auto-vectorization even makes the C version $\times400$
faster than the Python version (Note that to perform more reliable time
measurements, the size of the problem has been increased to 100001 instead of
101).

There are two main reasons for the poor performance of Python:

@@

    \item[dynamic binding] each variable access is made through a dictionary
        look-up that binds variable names to variable instances;

@@

In that context, it does not make sense to speak about vectorization: the
nature of an operation is unknown to the interpreter until its execution, and
data are not even contiguous in memory.

@@

Several approaches have been proposed to solve the performance issue of Python:
using native modules, relying on Ahead-Of-Time (AOT) compilation or
Just-In-Time (JIT) compilation.

A native module is a shared library whose interface make it callable from the
Python interpreter using the regular \texttt{import} mechanism. It makes it
possible to call native code from Python in a transparent way. Typically Python
object are unboxed when transfered from Python to C, where efficient
computations can happen. Then the result is boxed when getting back to the
Python world. In between, vectorization can happen. This is the approach taken
by the \texttt{numpy} module: a native type, the \texttt{ndarray} contains
contiguous unboxed elements, say single precision floating point numbers, and
provides a great deal of high-level functions to manipulate them. All the
computation-intensive part is done in pure C.  Vectorization can eventually
happen at this level, for instance when performing a point-to-point operation
like the sum of two arrays. However the granularity of the operations limits
the ratio of \texttt{load}/\texttt{store} versus operation. For instance, if
\texttt{a}, \texttt{b} and \texttt{c} are 1-D arrays, the sum \texttt{a + b +
c} is computed using two loops and an intermediate array. It means there are 4
load / stores and only two additions per loop iteration and vectorization is
unlikely to be profitable.

AOT compilation uses type inference or type annotation to statically compute
the type of all (like in Shedskin~\cite{shedskin2006},
Parakeet~\cite{parakeet2012}) or part of (like in Cython~\cite{cython2010} or
Numba~\cite{numba}) variables, effectively unboxing them to speed-up
computations. Using JIT compilation, type information is available at runtime
and the unboxing decision is based on profitability, e.g.\ according to a trace
analysis. This is the approach of the PyPy project~\cite{pypy2009}. In both
cases, generation of vector instruction is possible, but none of the cited
project uses it. Indeed automatic vectorization is a complex topic, and as
stated in~\cite{maleki2011}, many cases are not yet correctly supported even
for mainstream C/C++/FORTRAN compilers.

@@

Pythran~\cite{pythran2013, pyhpc2013} is also an AOT compiler, but it holds the
unique property of turning Python modules into C++11 \emph{meta-programs}. It
does not need any type annotations during translation and keeps the
polymorphism of original functions, respecting Python's duck typing. Its
compilation work-flow is described in Figure~\ref{fig:pythran-compiler}: Python
source is turned into generic C++11 code that is then instantiated for the
given types provided as extra annotations.  Like other AOT compilers, it
unboxes all variables for better performance, but also performs a wide variety
of optimizations such as lazy evaluation, constant folding, expression fusion
etc.  It supports core \texttt{numpy} constructs and most Python constructs, to
the notable exception of user classes. Instructions such as \texttt{eval},
\texttt{exec} or \texttt{getattr} are not supported either.

@@

\caption{Pythran compiler work-flow.}
\label{fig:pythran-compiler}
\end{figure}

This paper explores the vectorization of Python programs based on the idea that
many Python constructs already exhibit good vectorization opportunities. Thus
the goal of the compiler switches from automatic vectorization to vectorization
of high-level Python intrinsics. One of the difficulties of vectorizing Python
functions is linked to function polymorphism: vectorization in the context of
C++ meta-programs is first examined in Section~\ref{sec:meta-vectorization},
and Python constructs that can benefit from vectorization are then described in
Section~\ref{sec:python-semantic}. The approach is validated using the Pythran
compiler on several kernels presented in Section~\ref{sec:benchs}.

@@

\boostsimd{} is a high-level C++ library to program SIMD architectures. 
Designed as an \textit{Embedded Domain Specific Language}, \boostsimd{} 
provides both expressiveness and performance by using generic programming 
to handle vectorization. \boostsimd{} does not only provide a portable 
way but also enables the use of common programming idioms when designing 
SIMD-aware algorithms. Two components are provided:

@@

    \item[portability] Only SSE was supported, moving to AVX or another vector
        instruction set would have required a great deal of engineering;

    \item[completness] Many basic mathematic functions such as cos or sin
        were not directly available in the target instruction sets;

    \item[typing] Pythran manipulates meta-functions with parametric types,
        while a given instruction set only provides concretes type.

@@

Pythran uses function objects to represent any Python functions, either
user-functions or functions from the standard library. The former are
automatically generated from the user code analysis, and the latter are
hand-coded in a library called \texttt{pythonic++} provided with the compiler.
As an example, the Python function \texttt{def foo(x): return 3 * x} is compiled
into a code similar to the one in Listing~\ref{lst:simple-pythran-output}. The
actual is slightly more complicated to handle type inference across several
statements, but the structure remains valid.

@@

A critical point is that this representation perfectly matches the requirements
of \boostsimd, as presented in Listing~\ref{lst:square-cxx}. The \texttt{foo}
function can be used indifferently on scalar types and vector types. This paves
the way for automatic vectorization of high level idioms.

@@

However, a high-level language such as Python exhibits another level of
vectorization opportunity: implicit loops. Implicit loops can be found in many
places, but this paper focuses on \emph{list comprehension} and \emph{set
comprehension}, the \emph{map} and \emph{sum} intrinsics, and \emph{array
operations} through the \texttt{numpy} module.

@@

The \texttt{sum} intrinsic is a typical case of potential vectorization of an
implicit loop. Its prototype is: \texttt{sum(sequence[, start]) $\rightarrow$
value} and its behavior is equivalent to the Python code in
Listing~\ref{lst:sum-py}. Depending on the type of \texttt{sequence}, and also
whether one needs to strictly conforms to the IEEE Standard for Floating-Point
Arithmetic (IEEE 754), vectorization of this reduction is a valid application
of \boostsimd's \texttt{fold}.

@@

Condition~\ref{enu:same-types} is a prerequisite for static typing and efficient
unboxing. For instance, any code that does not match it will fail to compile
using Pythran or Shed-Skin. It is always satisfied for the \texttt{ndarray}
container as it wraps unboxed arrays. Condition~\ref{enu:scalar-types} can be
checked at compile time. Python supports two native scalar types candidates for
vectorization: \texttt{int} (64 bits integer) and \texttt{float}
(double-precision floating point numbers). The \texttt{numpy} module adds support for a
wide range of signed and unsigned integers as well as single, double or
quadruple precision float. Depending on the target architecture, only some of
them can fit into vector registers. Finally condition~\ref{enu:contiguous} also
is a static property of a container. It is verified for C++ alternative to
Python's \texttt{list}, \texttt{str} and \texttt{ndarrays}, but generally not
for \texttt{set} or \texttt{dict}. Condition \ref{enu:random-access} is not
matched by \emph{generator}s, the Python incarnation of co-routines.

@@

As function have to be generic but some containers do not match required
properties, it is necessary to select a vectorized implementation of fall back
to a generic one using Substitution Failure Is Not An Error
(SFINAE)~\cite{metaprogramming2002}. Vectorized versions are implement in this manner.

@@

\begin{lstlisting}[language=c++, label={lst:sum-cxx}, caption={Meta-implementation of the \texttt{sum} intrinsic in C++11.}]
template<class T, class V=long>
typename std::enable_if<
  is_vectorizable<T>::value,
  typename T::value_type
>::type
sum(T const& sequence, V value=0L)
{
  // E is the type of an element of T
  typedef typename T::value_type E;
  // vE is a vector type of vsize elements
  typedef typename simd::native<E,
    BOOST_SIMD_DEFAULT_EXTENSION> vE;
  static const size_t vsize =
    simd::meta::cardinal_of<vE>::value;

  // parallel reduction
  vE vector_value = simd::splat<vE>(E());
  const size_t n = sequence.size();
  const size_t bound = n / vsize * vsize;
  for(size_t i = 0; i < bound; i+= vsize)
    vp += sequence.load(i);
  // inner sum of the vector
  value += boost::simd::sum(vp);

@@

Deciding if a given function is vectorizable can be approximated through a
composition rules. First, some functions are known to be vectorizable. This is
generally the case for the addition, subtraction etc. Depending on the target
vector instruction set, some operations may or may not be available as vector
instructions, e.g.\ trigonometric functions, but they can be provided as
composite instructions, as done by \boostsimd. Some are just not available, as
quadruple precision float addition. Starting from this set of vectorizable
function, one can use the \texttt{;} as a closure binary relation to compute a
subset of all vectorizable functions. Said otherwise, any sequence of operation
that only involve vector operations is a vector operation and a function whose
body is a sequence of vector operation or vectorizable functions is a
vectorizable function itself.

The preceding statement is only valid if the body of the function only
references vector functions and formal parameters, as the function must be
valid for both scalar types and vector types. Note that this perfectly matches
the translation of polymorphic Python functions into C++ meta-functions, as
detailed in Listings~\ref{lst:square-python} and~\ref{lst:square-cxx}. The core
idea is that the \texttt{square} function can be called indifferently on
floating point values or on vector of floating point values. Through
(automatic) template instantiation, the C++ compiler generates the actual
scalar and vector version from the same meta-function. As Pythran always
translates Python's polymorphic function in that form, writing a vector version
of the \texttt{map} intrinsic becomes possible. It is slightly more technical
than for the \texttt{sum} intrinsic due to its variadic number of arguments,
but it relies on the same concepts.

@@

% Generator expression is somehow similar to list comprehension, to the notable
% exception that the value returned is an \emph{iterator adaptator} on the
% container argument. In the context of vectorization, a generator expression
% supports the vector protocol if the associated container also supports it. In
% that case the  \emph{iterator adaptator} implements the vector protocol by
% applying the comprehension rule to each vector element from this container, in a
% streaming fashion.

Python also provide same features for set and dictionary.  Set comprehension
can be rewritten in the form of generator expression using the eponym function:
\texttt{\{ x + 3 for x in l\}} is equivalent to \texttt{set(x + 3 for x in l)}.

Dictionary comprehension, generator expressions and multi-level comprehension
(i.e.\ containing multiple \texttt{for} clauses) are not considered in this
paper.

@@

The \texttt{numpy} module played a major role in the adoption of Python by the
scientific community. In a nutshell, it is a thin Python wrapper above a raw
array that provides an array language embedded into Python. An example of
\texttt{numpy} code is show on Listing~\ref{lst:numpy-code}.

@@

Common features of array language include indexing, slicing, point-to-point
operation etc. Point-to-point operations are traditionally a good source of
vectorization. However, because of the mandatory module layer, \texttt{numpy}
always create a new array to hold intermediate result. For instance in the
expression \texttt{(1 - x[:-1]) ** 2)} two array are created ---slicing does
not create an array but a view. The implied extra cost would hide the benefits
of vectorization.

The typical way to solve this issue in C++ is to use \emph{expression
templates}, a technique described in~\cite{expression_templates} that uses the
C++ type system to perform lazy evaluation of an expression. Yet, this
approach is limited to a single expression, as illustrated by Listing~\ref{lst:numpy-code-fs}: in that case expression templates would not prevent the creation
of temporary arrays \texttt{t0} and \texttt{t1}. Fortunately, it is possible to
apply \emph{forward substitution} to fall-back to the expression from
Listing~\ref{lst:numpy-code} and avoid the creation of any temporary in some
case detected with static analysis of the source code.

@@

It is possible to hook in the expression template code to add vectorization support.
Indeed the expression from Listing~\ref{lst:numpy-code} hides a single vector
loop, that can be explicitly vectorized using \boostsimd in a similar manner as
the \texttt{map} and \texttt{sum} intrinsics.

@@

\subsection{Vectorization of \texttt{ndarray} point-to-point operations with
intermediate array}

To have good performance even with intermediate array, Pythran uses forward
substitution. This is not always beneficial if the substituted variable is used multiple time ---it
would lead to the computation of the expression several times--- and it is only legal if no dependency are violated. This detection
transforms three SIMD loop computations in one intensive computation in the case of Listing~\ref{lst:numpy-code}. Figure~\ref{fig:numpy-code-fs} shows the benefits of the approach.

    \includegraphics[width=.5\textwidth]{rosen_fs}
    \caption{Timings of rosen computation with multiple assignment using CPython, Pythran, Pythran with vectorization, Pythran with vectorization and forward substitution.}
    \label{fig:numpy-code-fs}
